Species turnover reveals hidden effects of decreasing Nitrogen deposition in mountain hay meadows

### 2018-06-18 Manuscript.Rmd
Update Manuscript.Rmd

Since the late 1980s measures to reduce atmospheric pollution have successfully reduced emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition in many parts of Europe [@Torseth2012; @Fowler2007]. However, there are a number of factors that may prevent the recovery of plant communities that sufered from increased Nitrogen deposition. First, even if Nitrogen deposition is reduced, large areas might still be above the critical load (>>@Slootweg2015; @RihmAchermann<<). Second, there is a potential that communities reach an alternative stable state after decades of increased Nitrogen deposition and these species are unlikely to disappear again after the reduction of Nitrogen depositon [@Stevens2016]. Therefore, it is an open question whether and how fast the reduction in Nitrogen deposition rates will lead to the recovery of extant plant communities.

@@

In Switzerland, grassland accounts for 70 per cent of the agricultural land. With extensive cultivation, permanent grassland has a very high biodiversity. This applies in particular to the meadows in the alpine region, where meadows with a high plant diversity are of great importance agronomically (Leiber et al. 2005), for the image of the products and for the conservation of biodiversity (Hohl, 2006). Observation data of vascular plants from field recordings shows negative correlations of Nitrogen deposition with species richness and species composition in mountain hay meadows [@Roth2013]. Between 1990 and 2010 NOx emissions in Switzerland decreased by 46% and NH3 emissions by 14% (>>REF EINFÜGEN: Maas and Grennfelt, 2016<<), suggesting that Nitrogen deposition has decreased also in mountain hay meadows and we expected plant communities to respond accordingly.

@@

## Environmental gradients {-}
We expected different global change patterns to simultaneously cause temporal change in mountain hay communities. To better disentangle the importance of these mechanisms we ordered the sites along four main environmental gradients. First we expected communities to response to climate warming and the response might be different depending along the temperature gradient [@Roth2014]. To describe the temperature gradient we used the mean annual temperature per site from the WordClim database [@Fick2017]. Another key drivers for plants that is likely to be affected by climate change is precipitation [@Beier2012]. We used the annual precipitation per site from the WordClim database [@Fick2017]. Further, we estimated atmospheric Nitrogen deposition for each site using a pragmatic approach described in @Rihm2001 that combines monitoring data, spatial interpolation methods, emission inventories, statistical dispersion models and inferential deposition models. For details on the estimation of Nitrogen deposition see @Rihm2001 and @Roth2017. Finally we used inclination as proxy for land-use intensity because we expected that steeper sites are likely to be less intensively managed [@Strebel2015].

>>>>

Since the late 1980s measures to reduce atmospheric pollution have successfully reduced emissions of nitrogen oxides, with affiliated decrease in Nitrogen deposition in many parts of Europe [@Torseth2012; @Fowler2007]. However, there are a number of factors that may prevent the recovery of plant communities that sufered from increased Nitrogen deposition. First, even if Nitrogen deposition is reduced, large areas might still be above the critical load [@Slootweg2015; @RihmAchermann2016]. Second, there is a potential that communities reach an alternative stable state after decades of increased Nitrogen deposition and these species are unlikely to disappear again after the reduction of Nitrogen depositon [@Stevens2016]. Therefore, it is an open question whether and how fast the reduction in Nitrogen deposition rates will lead to the recovery of extant plant communities.

@@

In Switzerland, grassland accounts for 70 per cent of the agricultural land. With extensive cultivation, permanent grassland has a very high biodiversity. This applies in particular to the meadows in the alpine region, where meadows with a high plant diversity are of great importance agronomically (Leiber et al. 2005), for the image of the products and for the conservation of biodiversity (Hohl, 2006). Observation data of vascular plants from field recordings shows negative correlations of Nitrogen deposition with species richness and species composition in mountain hay meadows [@Roth2013]. Between 1990 and 2010 NOx emissions in Switzerland decreased by 46% and NH3 emissions by 14% [@MaasGrennfelt2016], suggesting that Nitrogen deposition has decreased also in mountain hay meadows and we expected plant communities to respond accordingly.

@@

## Environmental gradients {-}
We expected different global change patterns to simultaneously cause temporal change in mountain hay communities. To better disentangle the importance of these mechanisms we ordered the sites along four main environmental gradients. First we expected communities to response to climate warming and the response might be different depending along the temperature gradient [@Roth2014]. To describe the temperature gradient we used the mean annual temperature per site from the WordClim database [@Fick2017]. Another key drivers for plants that is likely to be affected by climate change is precipitation [@Beier2012]. We used the annual precipitation per site from the WordClim database [@Fick2017]. Further, we estimated atmospheric Nitrogen deposition for each site using a pragmatic approach described in @Rihm2001 that combines monitoring data, spatial interpolation methods, emission inventories, statistical dispersion models and inferential deposition models. For details on the estimation of Nitrogen deposition see @RihmAchermann2016 and @Roth2017. Finally we used inclination as proxy for land-use intensity because we expected that steeper sites are likely to be less intensively managed [@Strebel2015].
